Fail Faster

### 2025-03-16 paper/paper.tex
rough

 The locus for improvement in this area is \textit{generator libraries}: if programmers are 
 provided with the tools to write fast generators, then they do not themselves 
 need to be performance experts. However, existing libraries fail to achieve good performance 
 due to two major sources of inefficiency. The first is \textit{random sampling}, or the process
 of selecting random values. (Another sentence here?)
 The second is abstraction: generator libraries are implemented as embedded DSLS, which are flexible
 and expressive, but also introduce significant overhead that is not necessarily optimized away by 
 the compiler. These costs compound in the aggregate, raising a fundamental
 question: can generator libraries produce code that is both expressive \textit{and} efficient?

  % in strict functionallanguages like OCaml and Scala
We outline a fully general, two-stage approach to designing generator libraries, 
which provide high-level combinator functions with dramatically reduced 
overhead. First, \rand choice matters. We demonstrate that using a faster \rand
significantly impacts generator performance by directly optimizing an existing RNG, which
produces the same values as its unoptimized counterpart---but faster--thereby allowing us
to compare the RNG's impact on an existing generator library. Second, we 
describe a method of optimizing generator libraries using \textit{staged metaprogramming}, 
or ``staging''. Staged metaprogramming allows for code generation, which can mitigate 
performance challenges introduced by embedded DSLs. In particular, when a generator DSL 
introduces overhead for dynamic control flow, staging can specialize that control flow
and compile it statically. For example, when choosing between a list of options, a staged
generator can avoid allocating the list of options at runtime by expanding to an 
if-else. This approach is not tied to any particular PBT library---instead, it gives a recipe
for improving even SOTA PBT libraries.

To demonstrate
the effectiveness of our technique, we apply our technique to a leading PBT library, \bq, built with performance engineering in mind.
Our version of this library, termed \name, is a direct drop-in replacement for the original that is completely
semantically equivalent: generators written using it produce exactly the same values as 
those written in the original version. However, in \name, the abstractions provided by the generator DSL are 
zero-cost, and the \rand is highly-optimized. The direct equivalence between \name and \bq
allows us to perform extremely fine-grained comparisons between generators---generators implemented with both versions of the
library will produce exactly the same sequence of values, so differences in testing performance are directly attributable
to \name's performance speedup.

@@

We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
    namely, \rand choice, and abstraction overhead. We describe a general procedure to address the latter concern based on staged metaprogramming. \secref{section:motiv}
    \item We implement a version of \bq, \name, that contains our optimizations and maintains strict program equality. \secref{section:4}
    \item We evaluate \name by demonstrating the improved performance of generators constructed
    using our library in a controlled comparison, as well as its impact on bug-finding ability. \secref{section:eval}
\end{enumerate}

>>>>

The locus for improvement in this area is \textit{generator libraries}: if 
programmers are provided with the tools to write fast generators, then they 
do not themselves need to be performance experts. However, existing generator 
libraries fail to achieve good performance due to two major sources of 
inefficiency. The first is \textit{random sampling}, or the process of 
selecting random values, which, depending on one's choice of \rand and
the number of samples needed, can constitute an unexpectedly high proportion
of a generator's runtime. The second is \textit{abstraction}: while flexible 
and expressive, generator libraries rely on dynamic control structures that
allocate at runtime. Allocations compound in the aggregate, leading to costly
garbage collection pauses. This raises a fundamental 
question: can generator libraries produce code that is both expressive 
\textit{and} efficient?

  % in strict functionallanguages like OCaml and Scala
We outline a fully general, two-stage approach to designing generator 
libraries so that they retain their idiomatic abstractions with 
significantly improved performance. First, we demonstrate the impact of
RNG choice by optimizing an existing \rand. Our optimized RNG produces 
exactly 
the same values as its unoptimized counterpart, allowing us to directly 
measure the impact of this choice on generator performance. Second, 
we propose a method of streamlining generator libraries using 
\textit{staged metaprogramming}, or ``staging''. \tr{Joe---maybe add
a sentence here about parsing literature?} Using staging, it is possible
to eliminate the overhead of many common generator abstractions. For example,
where an unstaged generator DSL might represent a choice between
a list of options as a dynamically-allocated array of closures, a staged 
generator can expand to an if-else tree at compile 
time---eliminating the allocation entirely. This approach is not tied to any
particular PBT library, but rather provides a recipe for improving even
state-of-the-art implementations.

To demonstrate the effectiveness of our technique, we apply it to a leading PBT 
library,\\ \bq, built with performance engineering in mind.
Our version of this library, termed \name, is a drop-in replacement for \bq which 
is completely semantically equivalent: that is,  generators written using \name 
produce exactly the same values as those written in \bq. However, in \name, 
the abstractions provided by the generator DSL are zero-cost, and the \rand is 
highly-optimized. The direct equivalence between \name and \bq allows us to 
perform extremely fine-grained comparisons between generators, as generators 
implemented using both versions of the library will produce exactly the same 
sequence of values; so, differences in testing performance are directly 
attributable to \name's performance speedup.
\tr{Not really done... }

@@

In summary, we demonstrate that, across languages, monadic PBT generator DSLs can
have a significant performance overhead, and present a cross-language technique
for eliminating the overhead while preserving the idiomatic abstraction.
We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
    namely, \rand choice, and abstraction overhead. We describe a general procedure to address the latter concern based on staged metaprogramming. \secref{section:motiv}
    \item We implement a version of \bq, \name, that contains our optimizations and maintains strict program equality. \secref{section:4}
    \item We evaluate \name by demonstrating the improved performance of generators constructed
    using our library in a controlled comparison, as well as its impact on bug-finding ability. \secref{section:eval}
\end{enumerate}
